A cautionary note on the use of Ornstein Uhlenbeck and other models in macroevolutionary studies

### 2015-03-13 Manuscript/ohyou.tex
Started to re-write manuscript

\section{Abstract 100-200}
Phylogenetic comparative methods are increasingly used to give new insight into variation, causes and consequences of trait variation among species. The foundation of these methods is a suite of models that attempt to capture evolutionary patterns by extending the Brownian constant variance model. However, the parameters of these models have been hypothesised to be biased and only asymptotically behave in a statistically predictable way as datasets become large. This does not seem to be widely appreciated. We show that a commonly used model in evolutionary biology (the Ornstein-Uhlenbeck model) is biased over a wide range of conditions. Many studies fitting this model use datasets that are small and prone to substantial biases. Our results suggest that simulating fitted models and comparing with empirical results is critical when fitting OU and other extensions of the Brownian model.

@@

\section{Introduction}
\label{section:introduction} 
The outcomes of evolutionary processes acting over millions of years can be seen in the distributions and covariances of species traits at the tips of phylogenetic trees. Such processes can be modeled using phylogenetic comparative methods (PCMs). These approaches have been used, for example, to model niche conservatism \citep{Wiens:2010aa}, infer potential rates of species responses to climate change \citep{Quintero:2013aa}, test the role of ecological niche as a driver of morphological evolution \citep{pienaar2013macroevolution}, compare rates of evolution of species traits \citep{Claramunt:2012ab} and test for constraints in adaptive radiations \citep{blackburn2013adaptive}. PCMs are now widely recognized as powerful tools with which to model observed patterns in species traits and potentially to infer the evolutionary processes that drive them \citep[e.g.,][]{freckleton2009seven,Nunn:2011aa,o2012evolutionary,pennell2013integrative}. Despite these advances, the properties of some of the most widely used PCMs are poorly understood leading to the potential for inappropriate use and misinterpretation of results.

The majority of PCMs use an explicit evolutionary model to characterize trait evolution \citep{Freckleton:2011aa}. A model can be fit using statistical methods (typically maximum likelihood) and model parameters interrogated to test ideas from evolutionary theory. Most model-based methods for characterizing trait evolution are based on the Brownian constant variance model \citep[for exceptions see][]{price1997correlated,harvey2000comparative,freckleton2006detecting}. The Brownian model, first applied in a phylogenetic context by \citet{cavalli1967} and to across-species data by \citet{felsenstein1973maximum}, is a simple model of trait evolution in which trait variance accrues as a linear function of time, and makes the prediction that traits of closely-related species are more similar than those of distantly-related ones. The Brownian model has been modified in various ways to account for a suite of ecological and evolutionary processes (e.g., Grafen 1989; Hansen 1997; Pagel 1997, 1999). Most of these involve a transformation of the tree and thereby fitting a model with one or more extra parameters. These modified Brownian models have become very popular because they fit better, have attractive biological interpretations, and are implemented in freely available packages \citep[e.g., R;][]{R-Core-Team:2014aa}.

The modification of Brownian models to account for non-Brownian processes can be traced back to \citet{grafen1989phylogenetic} who introduced a parameter $\rho$ to account for non-linear scaling of evolutionary changes. $\rho$ is a power transformation of node heights in the phylogeny, and is fitted by maximum likelihood. It is worth noting that Grafen warned explicitly about potential drawbacks with $\rho$. Specifically: (i) the parameter is likely to be biased; (ii) its asymptotic sampling variance is non-zero; and (iii) values of the parameter should be interpreted cautiously. \citet{freckleton2002phylogenetic} showed that $\rho$ was indeed biased as predicted by Grafen (1989). They also showed that another parameter, Pagel’s $\lambda$ \citep{Pagel:1997aa,Pagel:1999aa}, also exhibited biases (but that these were likely to have minimal consequences in analyses using reasonable sample sizes). This strongly suggests that simulations are required to test the properties of such parameters, and that as recommended by \citet{grafen1989phylogenetic} “careless” parameterization of the phylogenetic variance-covariance matrix should be avoided.

@@

\section{Materials and Methods}
\subsection{Model Outline} 
\label{section:models} 
According to the Brownian model, a trait X evolves at random at a rate $\sigma$:

@@

@@

@@

\subsection{Literature Review}
\label{section:literature.methods} 
To get an overview of the use of OU models in ecology and evolution, we used Google Scholar (accessed 21st November 2013) to locate papers published between 2005 \citep[when the R package ouch was released;][]{Butler:2004aa} and 2013 that contained the terms “Ornstein Uhlenbeck” and either “ecology”, or “evolution” and “biology” (the “biology” term was added to omit physics papers which also use the term “evolution”). We also recorded the total number of papers containing the terms “ecology”, or “evolution” and “biology” published between 2005 and 2013 and plotted the number of OU papers published each year as a proportion of the total number of papers published (Fig. \ref{figure.literature}).

@@

\subsection{Literature Review}
\label{section:literature.results}
In total, 3720 papers containing the terms “Ornstein Uhlenbeck” and either “ecology”, or “evolution” and “biology” were published between 2005 and 2013, and the number has increased substantially since 2005 (Fig. \ref{figure:literature}). Most papers fit OU models to phylogenies with fewer than 100 taxa (mean = 166.97 $\pm$ 43.86, median = 58, Supplemental Fig. S1 and Table S1). The majority of papers fit OU models using R packages, particularly GEIGER and, although other uses are becoming more common, most papers use OU models in an effort to discern the “best” model of trait evolution or to control for phylogenetic non-independence (Supplemental Table S1 and Fig. S1).

>>>>

\section{Abstract 100-200}
%Phylogenetic comparative methods are increasingly used to give new insight into variation, causes and consequences of trait variation among species. The foundation of these methods is a suite of models that attempt to capture evolutionary patterns by extending the Brownian constant variance model. However, the parameters of these models have been hypothesised to be biased and only asymptotically behave in a statistically predictable way as datasets become large. This does not seem to be widely appreciated. We show that a commonly used model in evolutionary biology (the Ornstein-Uhlenbeck model) is biased over a wide range of conditions. Many studies fitting this model use datasets that are small and prone to substantial biases. Our results suggest that simulating fitted models and comparing with empirical results is critical when fitting OU and other extensions of the Brownian model.

@@

\section{Introduction}
\label{section:introduction} 
%The outcomes of evolutionary processes acting over millions of years can be seen in the distributions and covariances of species traits at the tips of phylogenetic trees. Such processes can be modeled using phylogenetic comparative methods (PCMs). These approaches have been used, for example, to model niche conservatism \citep{Wiens:2010aa}, infer potential rates of species responses to climate change \citep{Quintero:2013aa}, test the role of ecological niche as a driver of morphological evolution \citep{pienaar2013macroevolution}, compare rates of evolution of species traits \citep{Claramunt:2012ab} and test for constraints in adaptive radiations \citep{blackburn2013adaptive}. PCMs are now widely recognized as powerful tools with which to model observed patterns in species traits and potentially to infer the evolutionary processes that drive them \citep[e.g.,][]{freckleton2009seven,Nunn:2011aa,o2012evolutionary,pennell2013integrative}. Despite these advances, the properties of some of the most widely used PCMs are poorly understood leading to the potential for inappropriate use and misinterpretation of results.

%The majority of PCMs use an explicit evolutionary model to characterize trait evolution \citep{Freckleton:2011aa}. A model can be fit using statistical methods (typically maximum likelihood) and model parameters interrogated to test ideas from evolutionary theory. Most model-based methods for characterizing trait evolution are based on the Brownian constant variance model \citep[for exceptions see][]{price1997correlated,harvey2000comparative,freckleton2006detecting}. The Brownian model, first applied in a phylogenetic context by \citet{cavalli1967} and to across-species data by \citet{felsenstein1973maximum}, is a simple model of trait evolution in which trait variance accrues as a linear function of time, and makes the prediction that traits of closely-related species are more similar than those of distantly-related ones. The Brownian model has been modified in various ways to account for a suite of ecological and evolutionary processes (e.g., Grafen 1989; Hansen 1997; Pagel 1997, 1999). Most of these involve a transformation of the tree and thereby fitting a model with one or more extra parameters. These modified Brownian models have become very popular because they fit better, have attractive biological interpretations, and are implemented in freely available packages \citep[e.g., R;][]{R-Core-Team:2014aa}.

%The modification of Brownian models to account for non-Brownian processes can be traced back to \citet{grafen1989phylogenetic} who introduced a parameter $\rho$ to account for non-linear scaling of evolutionary changes. $\rho$ is a power transformation of node heights in the phylogeny, and is fitted by maximum likelihood. It is worth noting that Grafen warned explicitly about potential drawbacks with $\rho$. Specifically: (i) the parameter is likely to be biased; (ii) its asymptotic sampling variance is non-zero; and (iii) values of the parameter should be interpreted cautiously. \citet{freckleton2002phylogenetic} showed that $\rho$ was indeed biased as predicted by Grafen (1989). They also showed that another parameter, Pagel’s $\lambda$ \citep{Pagel:1997aa,Pagel:1999aa}, also exhibited biases (but that these were likely to have minimal consequences in analyses using reasonable sample sizes). This strongly suggests that simulations are required to test the properties of such parameters, and that as recommended by \citet{grafen1989phylogenetic} “careless” parameterization of the phylogenetic variance-covariance matrix should be avoided.

@@

\section{Materials and Methods}
According to the Brownian model \cite{cavalli1967,felsenstein1973maximum}, a trait X evolves at random at a rate $\sigma$:

@@

where $W(t)$ is drawn at random from a normal distribution with mean $0$ and variance $\sigma^2$. The model assumes that there is no overall drift in the direction of evolution (hence the expectation of $W(t)$ is zero) and that the rate of evolution is constant. Because the direction of change in trait values at each step is random, Brownian motion is often described as a ``random walk''. The model assumes the correlation structure among trait values is proportional to the extent of shared ancestry for pairs of species. This means that close relatives will be more similar in their trait values than more distant relatives. It also means that variance in the trait will increase (linearly) in proportion to time. The model has two parameters, the Brownian rate parameter, $\sigma^2$ and the state of the root at time zero, $X(0)$.

The Ornstein-Uhlenbeck (OU) model \citep{hansen1997stabilizing,Butler:2004aa} is a random walk where trait values revert back towards some ``optimal'' value with an attraction strength proportional to the parameter $\alpha$. The model has the following form:

Note that this model has two parameters in addition to those of the Brownian model: $\alpha$ and $\mu$. The parameter $\mu$ is a long-term mean, and it is assumed that species traits evolve around this value. $\alpha$ is the strength of evolutionary force that returns traits back towards the long-term mean if they evolve away from it. $\alpha$ is sometimes referred to as the ``rubber band'' parameter because of the way it pulls traits back towards $\mu$.

The OU model was introduced to population genetics by \cite{Lande:1976aa} to model stabilizing selection in which the mean was recast as a fitness optimum on an adaptive landscape. The process operating in comparative data is analogous, although clearly is not stabilizing selection, despite being sometimes referred to as such (see Discussion).

The variances and covariances predicted by the OU model are more complex than those predicted by the Brownian model. In the light of the results below, some properties of this model are worth highlighting:

  \item If $\alpha$ is small then evolution is approximately Brownian, The question is what is a small value of $\alpha$? Figure \ref{figure.traitsim} provides some guidance on this issue.\\

  \item In the long-term the imprint of history is weakened. If evolution proceeds for a long time the variance in $X_i$ tends to a constant, i.e. because the expected value of $X_i$ is $\mu$. Similarly, in equation (4), the covariance between traits tends to a constant because $T$ becomes large relative to $t_{ij}$. Consequently for large groups the model implies that the imprint of history is weak. As a result adding more species to a phylogeny will contribute only relatively weakly to fitting the model, as the deeper branches within a phylogeny will contribute little information to estimating the parameters.\\

Figure \ref{figure.traitsim} gives some numerical examples, chosen to outline the behavior of $\alpha$ in relation to the above. We show the predicted covariance in traits \ref{equation:OUcov} scaled by the overall variance \ref{equation:OUvar} plotted against the time since divergence, $t_{ij}$ scaled by tree height. Figure \ref{figure.traitsim} highlights that the behavior of $\alpha$ can be classified into three regions. When $\alpha$ is low the model behaves very much like Brownian motion, with the covariance scaling approximately linearly with time. Only as the value becomes moderately larger (in this case approaching a value of $0.5$), does the non-Brownian behavior become apparent. Secondly, there is a region of moderate values in which the model leads to markedly non-Brownian patterns of trait variation (Fig. \ref{figure.traitsim}B).

Finally, when $\alpha$ is large the model behaves differently again: in this case, all imprint of history is lost and the evolution is essentially a rapid burst at the present (Fig. \ref{figure.traitsim}C). This will be indistinguishable from phylogenetically unstructured data.

In terms of statistical model fitting, a feature to highlight is that in Figure \ref{figure.traitsim}A and Figure \ref{figure.traitsim}C, changing the value of α across large relative ranges at the extremes (0-0.5 in Fig. \ref{figure.traitsim}A; 10-50 in Fig. \ref{figure.traitsim}C) produces relatively little quantitative change in the model behavior, and it is in the middle region of parameter space (1-5) that the biggest discernible effects occur.

\section{Appendix}
\subsection{Model Outline} % correct the equations!
\label{section:models} 
According to the Brownian model, a trait X evolves at random at a rate $\sigma$:

  \begin{equation}
    dX(t) = \sigma dW(t)
    %\label{equation:BMrate} 
  \end{equation}

@@

  \begin{equation}
    dX(t) = - \alpha (X(t) - \mu) + \sigma dW(t)
    %\label{equation:OUrate} 
  \end{equation}

@@

\section{Supplementary Material}

  \subsection{Literature Review}
  \label{section:literature.methods} 
  To get an overview of the use of OU models in ecology and evolution, we used Google Scholar (accessed 13th March 2015) to locate papers published between 2005 \citep[when the R package ouch was released;][]{Butler:2004aa} and 2014 that contained the terms ``Ornstein Uhlenbeck'' and either ``'ecology'', or ``evolution'' and ``biology'' (the ``biology'' term was added to omit physics papers which also use the term ``evolution''). We also recorded the total number of papers containing the terms ``ecology'', or ``evolution'' and ``biology'' published between 2005 and 2014 and plotted the number of OU papers published each year as a proportion of the total number of papers published (Fig. \ref{figure.literature}).

@@

  \label{section:literature.results}
  In total, 3720 papers containing the terms “Ornstein Uhlenbeck” and either “ecology”, or “evolution” and “biology” were published between 2005 and 2013, and the number has increased substantially since 2005 (Fig. \ref{figure:literature}). Most papers fit OU models to phylogenies with fewer than 100 taxa (mean = 166.97 $\pm$ 43.86, median = 58, Supplemental Fig. S1 and Table S1). The majority of papers fit OU models using R packages, particularly GEIGER and, although other uses are becoming more common, most papers use OU models in an effort to discern the “best” model of trait evolution or to control for phylogenetic non-independence (Supplemental Table S1 and Fig. S1).
